% SIGPLAN template
\documentclass[preprint]{sigplanconf}

%% -- Packages Imports --

% AMS stuff
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{xspace}
\usepackage{comment}

% Language
\usepackage{csquotes}
\usepackage[british]{babel}
\MakeOuterQuote{"}


\newcommand{\name}{{\bf $\lambda_{\star}^{\mu}$}\xspace}
\newcommand{\coc}{{\bf $\lambda C$}\xspace}
\newcommand{\ecore}{$\lambda_{\star}$\xspace}
\newcommand{\cc}{$\lambda C$\xspace}
\newcommand{\sufcc}{{\bf Fun}\xspace}
\newcommand{\fold}{{\bf $\mathsf{fold}$}\xspace}
\newcommand{\unfold}{{\bf $\mathsf{unfold}$}\xspace}

\newcommand{\authornote}[3]{{\color{#2} {\sc #1}: #3}}
\newcommand\bruno[1]{\authornote{bruno}{red}{#1}}
\newcommand\jeremy[1]{\authornote{jeremy}{blue}{#1}}
\newcommand\linus[1]{\authornote{linus}{green}{#1}}
\newcommand{\fixme}[1]{{\color{red} #1}}

% Hyper links
\usepackage{hyperref}
\hypersetup{
   colorlinks,
   citecolor=black,
   filecolor=black,
   linkcolor=black,
   urlcolor=black
}

% Compact list
\usepackage{paralist}

% Figure import
\usepackage{graphicx}
\usepackage{float}

% Code highlighting
\usepackage{listings}
\usepackage{xcolor}
\newcommand{\hl}[2][gray!40]{\colorbox{#1}{#2}}
\newcommand{\hlmath}[2][gray!40]{%
  \colorbox{#1}{$\displaystyle#2$}}

% Theorem
\usepackage{amsthm}
\newtheorem{thm}{Theorem}[section]
\newtheorem{lem}[thm]{Lemma}
\newtheorem{dfn}[thm]{Definition}

%% Typesetting inference rules
\usepackage{styles/mathpartir}  % by Didier Rémy (http://gallium.inria.fr/~remy/latex/mathpartir.html

% Ott includes
\input{sections/expcore.ott.tex}
\input{sections/otthelper.tex}

% lhs2tex
\usepackage{mylhs2tex}

% Subsection style
\usepackage{titlesec}
\titleformat{\subsubsection}[runin]{\bfseries\itshape\normalsize}{}{0em}{}[$\;$]

% Table
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage[para,online,flushleft]{threeparttable}

%% -- Packages Imports --

% Main
\begin{document}

% Page size - US Letter
%\special{papersize=8.5in,11in}
%\setlength{\pdfpageheight}{\paperheight}
%\setlength{\pdfpagewidth}{\paperwidth}

% Conference info
\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country}
\copyrightyear{20yy}
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
\doi{nnnnnnn.nnnnnnn}

% Title
\titlebanner{} % Only for preprint
\preprintfooter{} % Only for preprint

\title{Type-Level Computation One Step at a Time}
%\title{A Dependently-typed Intermediate Language with General Recursion}
%\subtitle{Or: Decidable Type-Checking in the presence of
%  Type-Level General Recursion}

\authorinfo{}{}{}
% \authorinfo{Foo \and Bar \and Baz}
%            {The University of Foo}
%            {\{foo,bar,baz\}@foo.edu}

\maketitle

% Abstract
\begin{abstract}
Many type systems support a conversion rule that allows type-level
computation. In such type systems ensuring the \emph{decidability} of
type checking requires type-level computation to terminate.
For calculi where the syntax of types and terms is the same, the
decidability of type checking is usually dependent on the strong normalization
of the calculus, which ensures termination. An unfortunate
consequence of this coupling between decidability and strong
normalization is that adding (unrestricted) general recursion to such
calculi is difficult.

This paper proposes an alternative to the conversion rule that allows
the same syntax for types and terms, type-level computation, and
preserves decidability of type checking under the presence of general
recursion. The key idea, which is inspired by the traditional
treatment of \emph{iso-recursive types}, is to make each type-level
computation step explicit. Each beta reduction or expansion at the
type level is introduced by type-safe cast construct. Such construct allows control
over the type-level computation and ensures decidability of
type checking even in the presence of non-terminating programs at the
type level.  We realize this idea by presenting \name: a variant of the
calculus of constructions with general recursion and recursive types.
Furthermore we show how many advanced programming language features of
state-of-the-art functional languages (such as Haskell) can be encoded
in our minimalistic core calculus.
\end{abstract}

% Category, terms & keywords
\category{D.3.1}{Programming Languages}{Formal Definitions and Theory}
\terms Languages, Design
\keywords Dependent types, Intermediate language

%% -- Starting Point -- 

\input{sections/introduction.tex}

\input{sections/overview.tex}

\input{sections/examples.tex}

\input{sections/explicitcore.tex}

\input{sections/recursion.tex}

\input{sections/datatypes.tex}

\input{sections/related.tex}

\input{sections/conclusion.tex}

%% -- References --

% \acks
% Thanks to Blah. This work is supported by Blah.

\newpage

\bibliographystyle{abbrvnat}
% \nocite{*}
\bibliography{main}

%% -- Appendix --
\clearpage
\appendix
\input{sections/appendix.tex}

%% -- The end --

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
