\documentclass[oribibl]{llncs}
\usepackage{llncsdoc}
\usepackage{relsize}

\input{packages.tex}
\input{macros/pl-theory.tex}

\usepackage[numbers]{natbib}
\bibliographystyle{abbrvnat}
% \setcitestyle{aysep={}}

% Author notes
\newcommand{\authorNote}[3]{{\color{#2} {\textsc{#1}}: #3}}
% \newcommand\jeremy[1]{\authorNote{Jeremy}{pink}{#1}}
% \newcommand\bruno[1]{\authorNote{Bruno}{red}{#1}}
% \newcommand\tom[1]{\authorNote{Tom}{orange}{#1}}
\newcommand\name{$\lambda_{i}^{\!+}$\xspace}
\newcommand\oname{$\lambda_{i}$\xspace}
\newcommand\fname{$F_{i}$\xspace}
\newcommand\tname{$\lambda_{c}$\xspace}

% Ott includes
\inputott{ott-rules}
% I prefer rulenames on the right
\renewcommand\ottaltinferrule[4]{
  \inferrule*[narrower=0.7,right=#1,#2]
    {#3}
    {#4}
}

\renewcommand{\ottnt}[1]{#1}
\renewcommand{\ottmv}[1]{#1}


% Logical equivalence related macros
\newcommand{\valR}[2]{\mathcal{V}\bra{#1 ; #2}}
\newcommand{\valRR}[1]{\mathcal{V}\bra{#1}}
\newcommand{\eeR}[2]{\mathcal{E}\bra{#1 ; #2}}
\newcommand{\eeRR}[1]{\mathcal{E}\bra{#1}}
\newcommand{\ggR}[2]{\mathcal{G}\bra{#1 ; #2}}

\newcommand{\hl}[2][gray!40]{\colorbox{#1}{#2}}
\newcommand{\hlmath}[2][gray!40]{%
  \colorbox{#1}{$\displaystyle#2$}}


\begin{document}


\title{Coherent Nested Composition with Disjoint Intersection Types}

\ifdefined\submitoption
\subtitle{Long version with appendix}
\fi


\author{Xuan Bi\inst{1}  \and Bruno C. d. S. Oliveira\inst{1}  \and Tom Schrijvers\inst{2}}

\institute{
  The University of Hong Kong, \email{\{xbi,bruno\}@cs.hku.hk}
  \and
  KU Leuven, \email{tom.schrijvers@cs.kuleuven.be}
}

\maketitle


\begin{abstract}
Calculi with \emph{disjoint intersection types} support an introduction form
for intersections called the
\emph{merge operator}, while retaining a \emph{coherent} semantics.
% The interesting feature of such calculi is that they
%retain a coherent semantics, which is known to be hard in the presence
%of the merge operator. 
Disjoint intersections types have great potential to
serve as a foundation for powerful, flexible and yet type-safe and
easy to reason OO languages. This paper shows how to significantly
increase the expressive power of disjoint intersection types by
adding support for \emph{nested subtyping and composition}, which 
enables simple forms of \emph{family polymorphism} to be expressed in the calculus. 
%The motivation for those features is Ernst's \emph{family polymorphism}: the idea 
%that inheritance can be extended from a single class, to a whole
%family of classes. Nested subtyping and composition enable simple forms 
%of family polymorphism to be expressed, while retaining type-safety 
%and coherence. 
The extension with nested subtyping and
composition is challenging, for two different reasons. Firstly, the
subtyping relation that supports these features is non-trivial,
especially when it comes to obtaining an algorithmic version. Secondly,
the syntactic method used to prove coherence for previous calculi with disjoint
intersection types is too inflexible, making it hard to
extend those calculi with new features (such as nested subtyping).
We show how to address the first problem by adapting and extending
 the Barendregt, Coppo and Dezani (BCD) subtyping rules for intersections
with records and coercions. A sound and complete algorithmic
system is obtained by using an approach inspired by Pierce's
work. To address the second
problem we replace the syntactic method to prove coherence,
by a semantic proof method based on \emph{logical relations}.
Our work has been fully formalized in Coq, and we have an implementation
of our calculus.
\end{abstract}


% Main meat
\input{sections/introduction.tex}
\input{sections/overview.tex}
\input{sections/typesystem.mng}
\input{sections/coherence.mng}
\input{sections/algorithm.mng}
% \input{sections/discussion.tex}
\input{sections/related.tex}
\input{sections/conclusion.tex}


\newpage

\def\bibfont{\small}
\bibliography{paper}


\ifdefined\submitoption
\newpage
\appendix
\input{appendix.mng}
\fi


\end{document}
