\section{Introduction}
\label{sec:introduction}

Gradual typing~\citep{siek2006gradual} is an increasingly popular topic in both
programming language practice and theory. On the practical side there is a
growing number of programming languages adopting gradual typing. Those languages
include Clojure~\citep{Bonnaire_Sergeant_2016}, Python~\citep{Vitousek_2014, lehtosalo2016mypy},
TypeScript~\citep{typescript}, Hack~\citep{verlaguet2013facebook}, and the
addition of Dynamic to C\#~\citep{Bierman_2010}, to name a few. On the
theoretical side, recent years have seen a large body of research that defines
the foundations of gradual typing~\citep{garcia2016abstracting,
  cimini2016gradualizer, CiminiPOPL}, explores their use for both functional and
object-oriented programming~\citep{siek2006gradual, siek2007gradual}, as well as
its applications to many other areas~\citep{siek2016key, Ba_ados_Schwerter_2014,
  castagna2017gradual, Jafery:2017:SUR:3093333.3009865}.

A key concept in gradual type systems is
\emph{consistency}~\citep{siek2006gradual}. Consistency weakens type equality to allow
for the presence of \emph{unknown} types $[[ unknown ]]$. In some gradual type systems
with subtyping, consistency is combined with subtyping to give rise to
the notion of \emph{consistent subtyping}~\citep{siek2007gradual}. Consistent
subtyping is employed by gradual type systems to validate type
conversions arising from conventional subtyping. One nice feature of consistent
subtyping is that it is derivable from the more primitive
notions of \emph{consistency} % (arising from gradual typing)
and \emph{subtyping}. As \citet{siek2007gradual}
put it, this shows that
``gradual typing and subtyping are orthogonal and can be combined in a principled fashion''.
Thus consistent subtyping is often used as a guideline for
designing gradual type systems with subtyping. 

Unfortunately, as noted by \citet{garcia2016abstracting}, notions of
consistency and/or consistent subtyping ``become more difficult to adapt
  as type systems get more complex''. In particular, for the case of type
systems with subtyping, certain kinds of subtyping do not fit well with the
original definition of consistent subtyping by \citet{siek2007gradual}. One
important case where such a mismatch happens is in type systems supporting
implicit (higher-rank) polymorphism~\citep{jones2007practical,dunfield2013complete, odersky1996putting}.
It is well-known that polymorphic types
\`a la System F induce a subtyping relation that relates polymorphic types to
their instantiations~\citep{odersky1996putting, mitchell1990polymorphic}. However
\citeauthor{siek2007gradual}'s definition is not adequate for this kind of
subtyping. Moreover the current framework for \emph{Abstracting Gradual
Typing} (AGT)~\citep{garcia2016abstracting} does not account for
polymorphism either, but the authors
acknowledge that it is an interesting avenue for future work. 

Existing work on gradual type systems with polymorphism does not use
consistent subtyping. The Polymorphic Blame Calculus (\pbc)~\citep{ahmed2011blame, amal2017blame} is
an \emph{explicitly} polymorphic calculus with explicit casts, which is
often used as a target language for gradual type systems with
polymorphism. In \pbc a notion of \emph{compatibility} is employed
to validate conversions allowed by casts. Interestingly \emph{\pbc
  allows conversions from polymorphic types to their instantiations}.
For example, it is possible to cast a value with type
$[[ \/a. a -> a ]]$ into $[[ int -> int  ]]$. Thus an
important remark here is that, while \pbc is explicitly polymorphic,
casting and conversions are closer to \emph{implicit}
polymorphism. That is, in a conventional explicitly polymorphic calculus
(such as System F), the primary notion is type equality, where
instantiation is not taken into account. Thus the types
$[[ \/a. a -> a  ]]$ and $[[ int -> int  ]]$ are deemed
\emph{incompatible}. However in \emph{implicitly} polymorphic
calculi~\citep{jones2007practical,dunfield2013complete, odersky1996putting}
$[[ \/a. a -> a ]]$ and $[[ int -> int   ]]$ are deemed
\emph{compatible}, since the latter type is an instantiation of the
former.  Therefore \pbc is in a sense a hybrid between implicit and
explicit polymorphism, utilizing type equality (\`a la System F) for
validating applications, and \emph{compatibility} for validating
casts. 

An alternative approach to polymorphism has recently been proposed by
\citet{yuu2017poly}. Like \pbc their calculus is explicitly polymorphic.
However, they employ type consistency to validate cast conversions,
and forbid conversions from $[[ \/a. a -> a  ]]$ to $[[  int -> int  ]]$. This makes
their casts closer to explicit polymorphism, in contrast to \pbc. Nonetheless,
there is still some flavor of implicit polymorphism in their calculus when it
comes to interactions between dynamically typed and polymorphically typed code.
For example, in their calculus type consistency allows types such as $[[ \/a . a -> int  ]]$
to be related to $[[ unknown -> int  ]]$, where some sort of (implicit)
polymorphic subtyping is involved.


\begin{comment}
Instead several works use alternative notions such as
\emph{compatibility} of types~\citep{} or adapted versions of type
consistency~\citep{}.
For instance both \citet{ahmed2011blame} (in the
Polymorphic Blame calculus) and \citet{yuu2017poly} define
notions of \emph{compatibility}. \jeremy{inaccurate} These two notions of compatibility
are different from each other, and they are not derived from
consistency and subtyping.
\ningning{Mention however they are trying to mixing subtyping and consistency to some degree?}
\ningning{one advantage of their systems is that they do no guessing}
Thus one first criticism to compatibility is 
that it lacks the orthogonality of notions (subtyping and consistency)
afforded by consistent subtyping~\citep{siek2007gradual}  \bruno{Ningning, has this criticism
  been pointed out by someone else before? If so, who? Citing them
  would strengthen our point}.
\ningning{I think no except Siek}
Moreover, the 
proposals for compatibility are different, which makes it unclear 
of which one is more appropriate.
\end{comment}

The first goal of this paper is to study the gradually typed subtyping
and consistent subtyping relations for \emph{predicative implicit polymorphism}.
To accomplish this, we first show how
to reconcile consistent subtyping with polymorphism by generalizing
the original consistent subtyping definition by
\citeauthor{siek2007gradual}. Our new definition of consistent
subtyping can deal with polymorphism, and preserves the orthogonality
between consistency and subtyping. To slightly rephrase \citeauthor{siek2007gradual},
the motto of our paper is that:

\begin{quote}\emph{Gradual typing and {\bf polymorphism} are orthogonal and can be combined
in a principled fashion.}\footnote{Note here that we borrow \citeauthor{siek2007gradual}'s motto
  mostly to talk about the static semantics. As \citet{ahmed2011blame}
show there are several non-trivial interactions between
polymorphism and casts at the level of the dynamic semantics.}
\end{quote}

\noindent With the insights gained from our work, we argue that, for implicit
polymorphism, \citeauthor{ahmed2011blame}'s notion of compatibility is too
permissive (i.e., too many programs are allowed to type-check), and that
\citeauthor{yuu2017poly}'s notion of type consistency is too conservative. As a
step towards an algorithmic version of consistent subtyping, we present a
syntax-directed version of consistent subtyping that is sound and complete with
respect to our formal definition of consistent subtyping. The syntax-directed
version of consistent subtyping is remarkably simple and well-behaved, and does
not require the \emph{restriction} operator of \citet{siek2007gradual}.
Moreover, to further illustrate the generality of our consistent subtyping
definition, we show that it can also account for \emph{top types}, which cannot
be dealt with by \citeauthor{siek2007gradual}'s definition either.

%From the technical point of
%view the main innovation is a novel definition of consistent subtyping
%that generalizes Siek and Taha's, and preserves the orthogonality
%between subtyping and consistency. 

\begin{comment}
To further validate our revised notion of consistent subtyping, we
show that it coincides with the notion of consistent subtyping for an
extension of Garcia et al.'s \emph{Abstracting Gradual Typing} (AGT)
(2016) with polymorphism.
\end{comment}

The second goal of this paper is to present the design of \name, which
stands for \textbf{G}radually \textbf{P}olymorphic \textbf{C}alculus: a
(source-level) gradually typed calculus for (predicative) implicit higher-rank
polymorphism that uses our new notion of consistent subtyping. As far as we are
aware, there is no work on bridging the gap between implicit higher-rank
polymorphism and gradual typing, which is interesting for two reasons. On the
one hand, % from a practitioner's point of view,
modern functional languages (such as Haskell) employ sophisticated
type-inference algorithms that, aided by type annotations, can deal with
implicit higher-rank polymorphism. So a natural question is how gradual typing
can be integrated in such languages. On the other hand, there are several
existing works on integrating \emph{explicit} polymorphism into gradual typing
\citep{ahmed2011blame, yuu2017poly}. Yet no work investigates how to move its
expressive power into a source language with implicit polymorphism.
% In implicit polymorphism, type applications can be reconstructed by the type
% checker, for example, instead of \lstinline{id Int 3}, one can directly write
% \lstinline{id 3}.
Therefore as a step towards gradualizing such type systems, this paper develops
both declarative and algorithmic versions for a gradual type system with
implicit higher-rank polymorphism. The new calculus brings the expressive power
of full implicit higher-rank polymorphism into a gradually typed source language.
% Implicit polymorphism is a hallmark of functional programming, and
% modern functional languages (such as Haskell) employ sophisticated
% type-inference algorithms that, aided by type annotations, can deal with
% higher-rank polymorphism. Therefore as a step towards gradualizing such type
% systems, this paper develops both declarative and algorithmic versions for a
% type system with implicit higher-rank polymorphism.
% The algorithmic version
% employs techniques developed by \citet{dunfield2013complete} to deal with
% instantiation.
We prove that our calculus satisfies all of the \emph{static} aspects of the
refined criteria for gradual typing~\cite{siek2015refined}.

As a step towards the \emph{dynamic gradual
  guarantee} property~\citep{siek2015refined}, we propose an extension of our
calculus. This extension employs \emph{static type parameters},
which are placeholders for monotypes, and \emph{gradual type
  parameters}, which are placeholders for monotypes that are
consistent with the unknown type. The concept of static type
parameters and gradual type parameters in the context of gradual
typing was first introduced by \citet{garcia2015principal}, and later
played a central role in the work of \citet{yuu2017poly}\footnote{The
  static and gradual type variables in their work.}. With this
extension it becomes possible to talk about \emph{representative
  translations}: those translations that generalize a number of other
translations using specific monotypes. Our work recasts the dynamic gradual guarantee in terms of
representative translations. Assuming a coherence conjecture regarding
representative translations, the dynamic gradual guarantee of our
extended source
language now can be reduced to that of \pbc, which, at the time of writing, is
still an open question. Nonetheless, we believe our discussion of
representative translations is helpful in shedding some light on this issue.

In summary, the contributions of this paper are:
\begin{itemize}
\item We define a framework for consistent subtyping with:
  \begin{itemize}
  \item a new definition of consistent subtyping that subsumes and generalizes that
    of \citeauthor{siek2007gradual}, and can deal with polymorphism and top types, and
    % Unlike previous definitions of
    % \emph{compatibility} in \emph{Polymorphic Blame Calculus}
    % \citep{ahmed2011blame}, we show that the orthogonality between
    % subtyping and consistency can be preserved.
  \item a syntax-directed version of consistent subtyping that is
    sound and complete with respect to our definition of consistent
    subtyping, but still guesses instantiations.
  \end{itemize}
\item Based on consistent subtyping, we present \name: a declarative gradual type
  system with predicative implicit higher-rank polymorphism. We prove that our calculus
  satisfies the static aspects of the refined criteria for gradual
  typing~\citep{siek2015refined}, and is type-safe by a type-directed
  translation to \pbc\citep{ahmed2011blame}.
\item We present a sound and complete bidirectional algorithm for implementing
  the declarative system based on the design principle of
  \citet{garcia2015principal} and the approach of \citet{dunfield2013complete}.
  A Haskell implementation of the type checker is also available.
\item We propose an extension of the type system with type
  parameters~\citep{garcia2015principal} as a step towards restoring the
  \emph{dynamic gradual guarantee}~\citep{siek2015refined}. The extension
  significantly changes the algorithmic system. The new algorithm features a novel
  use of existential variables with a different solution space, which is a natural extension
  of the approach by \citet{dunfield2013complete}.
\item All of the metatheory of this paper, except some manual proofs for the
  algorithmic type system and extensions, has been mechanically formalized in
  Coq.
\end{itemize}


\tikzset{
  state/.style={
    rectangle,
    rounded corners,
    minimum height=2em,
    inner sep=2pt,
    text centered,
    align=center
  },
}

\begin{figure}[t]
\begin{tikzpicture}[->,>=stealth']
  \footnotesize

  \node[state,
  draw=black, very thick,
  % text width=2.3cm
  ](cons-and-sub)
 {\begin{tabular}{ll}
    \parbox{2.3cm}{\begin{itemize}
      \item Subtyping
      \item Consistency
      \end{itemize}
    }
    & (Fig.~\ref{fig:decl:subtyping})\\
  \end{tabular}
  };

 \node[state,    	% layout (defined above)
  % text width=3cm, 	% max text width
  below of=cons-and-sub, 	% Position is to the right of
  node distance=2cm, 	% distance
  anchor=center] (conssub) 	% posistion relative to the center of the 'box'
 {
   \parbox{1.4cm}{Consistent\\
     Subtyping (Def.~\ref{def:decl-conssub})}
 };

 \node[state,
  right of=conssub,
  anchor=center,
  node distance=3.8cm,
  ] (syntax-conssub)
 {
   \parbox{1.7cm}{\centering
     $[[dd |- A <~ B]]$\\
     (Fig.~\ref{fig:decl:conssub})}
 };

 \node[state,
  right of=syntax-conssub,
  anchor=center,
  node distance=3.2cm,
  ] (decl)
 {
   \parbox{1.4cm}{\centering
     $[[dd |- e : A]]$\\
     (Fig.~\ref{fig:decl-typing})}
 };

 \node[state,
  right of=decl,
  anchor=center,
  node distance=3.5cm,
  ] (algo)
 {
   \parbox{2cm}{\centering
     $[[ GG |- ae => aA -| DD ]]$\\
     $[[ GG |- ae <= aA -| DD ]]$\\
     (Fig.~\ref{fig:algo:typing})}
 };

 \node[state,
  below of=conssub,
  anchor=center,
  node distance=2.5cm,
  ] (siek)
 {
   \parbox{2cm}{\citeauthor{siek2007gradual}
     (Def.~\ref{def:old-decl-conssub})}
 };

 \node[state,
  below of=syntax-conssub,
  anchor=center,
  node distance=2.5cm,
 ] (agt)
 {
   \parbox{0.6cm}{AGT}
 };

 \node[state,
  above of=decl,
  anchor=center,
  node distance=2cm,
  ] (ol)
 {
   \parbox{2.5cm}{\centering
     $[[ dd ||- e : A ]]$\\
     (Fig.~\ref{fig:original-typing})}
 };

 \node[state,
  below of=decl,
  anchor=center,
  node distance=2.5cm,
  ] (pbc)
 {
   \parbox{1.2cm}{$[[ dd |- pe : A]]$}
 };

 % draw the paths and and print some Text below/above the graph
 \path
 (cons-and-sub)   edge   node[right]{superimposed}                                      (conssub)
 (conssub)        edge
     node[above]{Thm.~\ref{lemma:properties-conssub}}                                 (syntax-conssub)
 (syntax-conssub) edge                                                                  (conssub)
 (decl)           edge   node[above]{based on}                                          (syntax-conssub)
 ([yshift=0.2cm]decl.east) edge
     node[above]{Thm.~\ref{thm:type_complete}}                               ([yshift=0.2cm]algo.west)
 (algo)           edge   node[below]{Thm.~\ref{thm:type_sound}}                 (decl)
 (conssub)        edge
     node[right, text width=1.5cm]{generalizes (Prop.~\ref{prop:subsumes})}              (siek)
 (agt)            edge
     node[left, text width=1.5cm]{verified on simple types (Prop.~\ref{lemma:coincide-agt})}         (syntax-conssub)
 (agt)            edge[bend right]
     node[right, text width=2.5cm]{verified on $\tope$ (Prop.~\ref{prop:agt-top})}   (syntax-conssub)
 (decl)           edge   node[right]{gradualizes}                                        (ol)
 (decl)           edge
     node[right, text width=2cm]{type-directed translation (Thm.~\ref{lemma:type-safety})} (pbc)
     ;

\end{tikzpicture}
\caption{Some key results in the paper.}
\label{fig:key-results}
\end{figure}




\Cref{fig:key-results} summarizes some key results in the paper.  The reader is
advised to refer back to this figure while reading the rest of the paper, as what it
depicts will gradually come to make sense.

This article is a significantly expanded version of a conference
paper~\citep{xie2018consistent}. Besides many improvements and expansions of
the conference paper materials, there are several novel extensions. Firstly, we
have added \lstinline{let} expressions to our gradually typed calculus. Therefore
it can now be seen as a complete gradual
version of the implicitly polymorphic lambda calculus by \citet{odersky1996putting}.
Secondly, we have
significantly expanded the discussion of applications of the calculus. In
particular we now show how we can model algebraic datatypes, and how the
combination of gradual typing and polymorphism enables simple
(gradual) formulations
of heterogeneous data structures~\citep{kiselyov2004strongly,
  mcbride2002faking}. Thirdly, we provide an extensive discussion of a
variant of the calculus with a subsumption rule and prove its soundness and
completeness. % Fourthly, we now formally discuss more properties of the calculus
Fourthly, we study an extension of the calculus with type parameters and
discuss the \emph{dynamic gradual guarantee}~\citep{siek2015refined} in detail. Furthermore we 
formalize the notion of \emph{coherence up to cast errors} in Section~\ref{sec:type-system}.
We also include detailed proofs on decidability of the algorithmic system.
Finally, we
provide an implementation of our type checking algorithm.

All supplementary materials, including Coq mechanization, manual proofs, and the
Haskell implementation of the algorithm, are available at
\url{https://github.com/xnning/Consistent-Subtyping-for-All}.


% The rest of the paper is organized as follows. In \Cref{sec:background}
% we review a simple gradually typed language with objects, to introduce the
% concept of consistent subtyping. We also briefly review our original language:
% the Odersky-L{\"a}ufer type system for higher-rank types. We motivate and
% justify our new definition of consistent subtyping in
% \Cref{sec:exploration}. In \Cref{sec:type-system} we present the
% declarative type system for our gradually typed calculus with implicit
% higher-rank polymorphism. In \Cref{sec:algorithm} we give a simple
% algorithm to implement the declarative system, and show it is sound and complete
% with respect to the declarative system in \Cref{sec:sound:complete}. We
% then show a simple extension and discuss some issues in
% \Cref{sec:discussion}. After discussing related work in
% \Cref{sec:related}, we conclude in \Cref{sec:conclusion}.




%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../paper"
%%% org-ref-default-bibliography: "../paper.bib"
%%% End:
