% Springer template
\documentclass[a4paper]{llncs}

%% -- Packages Imports --

% AMS stuff
\usepackage{amsmath}
\usepackage{amssymb}
\let\proof\relax
\let\endproof\relax
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{xspace}
\usepackage{setspace}
\usepackage{comment}
\usepackage{cite}

% Language
\usepackage{csquotes}
\usepackage[british]{babel}
\MakeOuterQuote{"}


\newcommand{\name}{{\bf $\lambda_{\star}^{\mu}$}\xspace}
\newcommand{\coc}{{\bf $\lambda C$}\xspace}
\newcommand{\ecore}{$\lambda_{\star}$\xspace}
\newcommand{\cc}{$\lambda C$\xspace}
\newcommand{\sufcc}{{\bf Fun}\xspace}
\newcommand{\fold}{{\bf $\mathsf{fold}$}\xspace}
\newcommand{\unfold}{{\bf $\mathsf{unfold}$}\xspace}

\newcommand{\authornote}[3]{{\color{#2} {\sc #1}: #3}}
\newcommand\bruno[1]{\authornote{bruno}{red}{#1}}
\newcommand\jeremy[1]{\authornote{jeremy}{blue}{#1}}
\newcommand\linus[1]{\authornote{linus}{green}{#1}}
\newcommand{\fixme}[1]{{\color{red} #1}}

% Hyper links
\usepackage{hyperref}
\hypersetup{
   colorlinks,
   citecolor=black,
   filecolor=black,
   linkcolor=black,
   urlcolor=black
}

% Compact list
\usepackage{paralist}

% Figure import
\usepackage{graphicx}
\usepackage{float}

% Code highlighting
\usepackage{listings}
\usepackage{xcolor}
\newcommand{\hl}[2][gray!40]{\colorbox{#1}{#2}}
\newcommand{\hlmath}[2][gray!40]{%
  \colorbox{#1}{$\displaystyle#2$}}

% Theorem
% \usepackage{amsthm}
% \newtheorem{thm}{Theorem}[section]
% \newtheorem{lem}[thm]{Lemma}
% \newtheorem{dfn}[thm]{Definition}

%% Typesetting inference rules
\usepackage{styles/mathpartir}  % by Didier Rémy (http://gallium.inria.fr/~remy/latex/mathpartir.html

% Ott includes
\input{sections/expcore.ott.tex}
\input{sections/otthelper.tex}

% lhs2tex
\usepackage{mylhs2tex}

% Subsection style
% \usepackage{titlesec}
% \titleformat{\subsubsection}[runin]{\bfseries\itshape\normalsize}{}{0em}{}[$\;$]

% Table
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage[para,online,flushleft]{threeparttable}

%% -- Packages Imports --

% Main
\begin{document}

\mainmatter

% Title
\title{A Simple Yet Expressive Calculus for Modern Functional Languages}
%\title{A Dependently-typed Intermediate Language with General Recursion}
%\subtitle{Or: Decidable Type-Checking in the presence of
%  Type-Level General Recursion}

\author{Bi Xuan \and Yang Yanpeng \and Bruno C. d. S. Oliveira}

\institute{The University of Hong Kong\\
\email{\{xbihku, yangyp\}@connect.hku.hk} \\
\email{bruno@cs.hku.hk}}

% \authorinfo{Foo \and Bar \and Baz}
%            {The University of Foo}
%            {\{foo,bar,baz\}@foo.edu}

\maketitle

% Abstract
\begin{abstract}
Typed core (or intermediate) languages for modern 
functional languages, such as Haskell or ML, 
are becoming more and more complex. This is a natural tendency.
Programmers and language designers wish for more expressive and
powerful source-language constructs. In turn this requires new, more
powerful constructs in core languages. Unfortunately, the added
complexity means that the meta-theory and implementation of such core
languages becomes significantly harder.

This paper proposes a simple yet expressive core calculus (\name),
which has a fraction of the language constructs of existing core
languages. The key to simplicity is the combination of two ideas. The
first idea is to use a Pure Type Systems (PTS) style of syntax that
unifies the various syntactic levels of the language. However, this
creates an immediate challenge: with types and terms unified, the
\emph{decidability} of type checking requires type-level computation
to terminate, but with general recursion it is hard to have such
guarantee. The second idea, inspired by the traditional treatment of
iso-recursive types, is to solve this challenge by making each
type-level computation step explicit. The usefulness of \name is
illustrated by a light surface language built on top of \name, which
supports many advanced programming language features of
state-of-the-art functional languages. 
%The main limitation of \name is
%the absence of a more expressive form of type-equality, which is left
%for future work.
\end{abstract}

% Category, terms & keywords
% \category{D.3.1}{Programming Languages}{Formal Definitions and Theory}
% \terms Languages, Design
% \keywords Dependent types, Intermediate language

%% -- Starting Point -- 

\input{sections/introduction.tex}

\input{sections/overview.tex}

\input{sections/examples.tex}

\input{sections/explicitcore.tex}

% \input{sections/recursion.tex}

\input{sections/datatypes.tex}

\input{sections/related.tex}

\input{sections/conclusion.tex}

%% -- References --

% \acks
% Thanks to Blah. This work is supported by Blah.

\newpage

\bibliographystyle{plain}
% \nocite{*}
\bibliography{main}

%% -- Appendix --
\clearpage
\appendix
\input{sections/appendix.tex}

%% -- The end --

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
