\section{Tail-call Elimination}\label{sec:tce}

In this section, we show how we can augment the basic translation in
Section~\ref{sec:fcore} to support tail-call elimination.
 
As shown in Figure \ref{fig:even}, we can do TCE
with IFOs. To capture this formally, we augment the
\lstinline{apply} method call generation, in rule \texttt{CJ-App}, with two possibilities:

\begin{enumerate}

\item The \lstinline{apply} method is in a tail position. This means we can
  immediately return by setting the \texttt{next} field of the
  controlling auxiliary \lstinline{Next} class to the current \lstinline{Function} object,
  without calling the apply method.

\item The \lstinline{apply} method is not in a tail position. This means we
  need to evaluate the corresponding chain of calls, starting with the
  current call, followed by any apply calls within it. 

\end{enumerate}

We need to make two changes to achieve this goal: 1) add a tail call
detection mechanism; and 2) use a different way of compiling function
applications.

\paragraph{Detecting Tail Calls.}
We base the detection mechanism on the tail call context from the
Revised Report on Scheme \cite{Abelson1998}. When we translate a value
application $e_1~e_2$, we know that $e_2$ is not in a tail position,
whereas $e_1$ may be if the current context is a tail context.  In
type applications and abstractions, we know they only affect types: they do not
affect the tail call context.  Thus, they preserve the state we
entered with for translating the apply calls. In $\lambda$
abstractions, we enter a new tail call context. This detection mechanism 
is integrated in our translation and used when compiling function applications.

\paragraph{Compiling Function Applications.}
We augment the \lstinline{apply} method call generation as follows. We extend the 
premise of \texttt{CJ-App} to include one extra freshly generated variable $c$: 

{\small
 \bda{l}
  \ba{l}
\myirule{
\Gamma \turns e_1 : \type_2 \rightarrow \type_1
 \leadsto J_1 \textbf{~in~} \Gamma_{J1};C_1;S_1 ~~~~~~~\\
 \Gamma \turns e_2 : \type_2 \leadsto J_2 \textbf{~in~} \Gamma_{J2};C_2;S_2 ~~~~~~~
 f,~x_f~,~c~fresh
 }{
 \Gamma \turns e_1 \, e_2 : \type_1 \leadsto x_f \textbf{~in~} 
 \Gamma_{J1} \uplus \Gamma_{J2} \\ \uplus \{f : Function, c : Function,
 x_f : \langle\type_1\rangle\}; C_1, C_2;\{S_1;S_2;S_3\}
 }
  \ea
 \eda
}

In the conclusion, we change $S_3$. For tail calls, we define it as follows:

\begin{lstlisting}[mathescape]
$S_3$ :=  {
       Function f = $J_1$;
       f.arg = $J_2$;
       Next.next = f;
}
\end{lstlisting}

Note that $x_f$ is not bound in $S_3$ here. Because the result of a
tail call is delayed, the result of the tail call is still not
available at this point.  However, this does not matter: since we are
on a tail-call, the variable would be immediately out of its scope
anyway and cannot be used.

For non-tail calls, we initialize $x_f$ in $S_3$ as the final result:

\begin{lstlisting}[mathescape]
$S_3$ :=  {
       Function f = $J_1$;
       f.arg = $J_2$;
       Next.next = f;
       Function c;
       Object $x_f$;
       do {
         c = Next.next;
         Next.next = null;
         c.apply();
       } while (Next.next != null);
       $x_f$ = c.res;
}
\end{lstlisting}

This generated code resembles the example in
Section~\ref{sec:overview}, except for the general \lstinline{Object}
$x_f$ being in place of the specialized \lstinline{Boolean res}. The
idea of looping through a chain of function calls remains the same.
Note that the rules here do not produce valid MJ programs, because MJ does not support
\lstinline{static} fields and looping constructs. For restating the translation properties,
we would need to use a different formalization of Java that supports both of these features.
The \emph{type preservation} proof would be augmented in two ways: the global class table
$\Delta_G$ would also contain information about the \lstinline{Next} class and 
the value application case would have two sub-cases for tail and non-tail calls.
The \emph{semantic preservation} proof would require formalization of the tail call
detection mechanism and be even more involved due to the interaction between
tail and non-tail calls. Since this is a major endeavor, which is not in-scope
for this paper, we leave formalizing an extended subset of Java and 
proving the properties of the TCE-augmented translation for future work.