\section{Related Work}

This section discusses related work: intermediate functional languages on top of the JVM,
TCE and function representations, TCE on the JVM, and the JVM modifications.

\paragraph{Intermediate Functional Languages on top of the JVM.}
A primary objective of our work is to create an efficient intermediate
language that targets the JVM. With such intermediate language,
compiler writers can easily develop FP compilers in the JVM.
System F is an obvious candidate for an intermediate language as it
serves as a foundation for ML-style or Haskell-style FP languages.
However, there is no efficient implementation of System F in the JVM.
The only implementation of System F that we know of (for a JVM-like
platform) was done by Kennedy and Syme~\cite{Kennedy2004}. They showed
that System F can be encoded, in a type-preserving way, into
.NET's C\#. That encoding could easily be employed in Java or the JVM as
well. However, their focus was different from ours. They were not aiming
at having an efficient implementation of System F. Instead, their goal
was to show that the type system of languages such as C\# or Java is
expressive enough to faithfully encode System F terms. They used a
FAO-based approach and have not exploited the erasure semantics of System F.
As a result, the encoding suffers from various performance drawbacks
and cannot be realistically used as an intermediate language. MLj
\cite{Benton1998} compiled a subset of SML '97 (interoperable with
Java libraries) to the Monadic Intermediate Language, from which it
generated Java bytecode. Various Haskell-to-JVM compiler backends
\cite{Wakeling1999,Tullsen1996,Choi2001} used different
variations of the \emph{graph reduction machine}~\cite{Wadsworth:1971} for their
code generation, whereas we translate from System F.  Nu \cite{Dyer2008}
provided a similar intermediate layer targeting the JVM for developing
Aspect Oriented languages.

\paragraph{Tail-Call Elimination and Function Representations.}
A choice of a function representation plays a great role
\cite{Shao1994} in time and space efficiency as well as in how
difficult it is to correctly implement tail calls. Since
Steele's pioneering work on tail calls \cite{Steele1977}, implementors
of FP languages often recognize TCE as a necessary
feature. Steele's Rabbit Scheme compiler~\cite{Steele1978} introduced the ``UUO handler'' 
that inspired our TCE technique using IFOs.
Early on, some Scheme compilers targeted C as an intermediate
language and overcame the absence of TCE in the backend compiler by
using trampolines. Trampolines incur on performance penalties and
different techniques, with ``Cheney on the M.T.A." \cite{Baker1995}
being the most known one, improved upon them. The limitations of the
JVM architecture, such as the lack of control over the memory
allocation process, prevent a full implementation of Baker's
technique. 

\paragraph{Tail-Call Elimination on the JVM.}
Apart from the recent languages, such as Scala \cite{Odersky2014b} or
Clojure \cite{Hickey2008}, functional languages have
targeted the JVM since its early versions. 
%The experimental language Mondrian \cite{Perry2004} demonstrated
%different challenges of implementing functional languages on top of
%object-oriented virtual machine environments, such as .NET or
%JVM. Challenges include the question how to represent functions, but
%most languages use the standard FAO encoding.
Several other JVM functional languages support (self) tail recursion
optimization, but not full TCE. Examples include MLj \cite{Benton1998}
or Frege~\cite{Wechsung}.  Later work \cite{Minamide2003} extended MLj
with Selective TCE. This work used an effect system to estimate the number
of successive tail calls and introduced trampolines only when
necessary. Another approach to TCE in the JVM  is to 
use an explicit stack on the heap (an \texttt{Object[]} array)~\cite{Choi2001}. 
With such explicit stack for TCE, the approach from Steele's pioneering work \cite{Steele1978} 
can also be encoded in the JVM. Our work avoids the need for an explicit stack by using 
IFOs, thus allowing for a more direct implementation of this technique. 
%\cite{Steele1978}: a while loop runs function
%applications and when a function wants to make a tail call, it returns
%(or sets) a function to be executed next. 
%%Frege \cite{Wechsung} is a
%%functional language in the spirit of Haskell on JVM and avoids some
%%problems with tail calls through laziness. Similarly to our
%%implementation, it first compiles to Java expressions rather to
%%directly to the Java bytecode.
The Funnel compiler for the JVM \cite{Schinz2001} used
standard method calls and shrank the stack only after the execution
reached a predefined ``tail call limit". This dynamic optimization
needs careful tuning of the parameters, but can be possibly used to
further improve performance of our approach.

\paragraph{JVM Modifications.}
Proposals to modify the JVM \cite{League2001}, which would arguably be a better
solution for improving support for FP,
appeared early on. One reason why the JVM does not support tail calls
was due to a claimed incompatibility of a security mechanism based on
stack inspection with a global TCE policy. The abstract
continuation-marks machine \cite{Clements2004} refuted this
claim. There exists one modified Java HotSpot\texttrademark VM
\cite{Schwaighofer2009} with TCE support. The research Maxine VM
with its new self-optimizing runtime system \cite{Wuerthinger2013}
allows a more efficient execution of JVM-hosted languages. Despite
these and other proposals and JVM implementations, such as IBM J9, we
are not aware of any concrete plans for adding TCE support to the
next official JVM release. Some other virtual machines designed for
imperative languages do not support TCE either. For example, the
standard Python interpreter lacks it, even though some enhanced
variants can overcome this issue \cite{Tismer2000}. Hence, ideas from our
work can be applied outside of the JVM ecosystem.
