
\documentclass[a4paper,UKenglish]{lipics-v2018}
%This is a template for producing LIPIcs articles. 
%See lipics-manual.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
% for section-numbered lemmas etc., use "numberwithinsect"


\nolinenumbers

\usepackage{microtype}%if unwanted, comment out or use option "draft"

\input{packages.tex}
\input{macros/pl-theory.tex}

\newcommand{\name}{\textsf{NeColus}\xspace}
\newcommand{\namee}{$\lambda_{i}^{+}$\xspace}
\newcommand\oname{$\lambda_{i}$\xspace}
\newcommand\fname{$F_{i}$\xspace}
\newcommand\tname{$\lambda_{c}$\xspace}

% Ott includes
\inputott{ott-rules}
% I prefer rulenames on the right
% \renewcommand\ottaltinferrule[4]{
%   \inferrule*[narrower=0.7,right=#1,#2]
%     {#3}
%     {#4}
% }
\renewcommand\ottaltinferrule[4]{
  \inferrule*[narrower=0.5,lab=#1,#2]
    {#3}
    {#4}
}

% \renewcommand{\ottnt}[1]{#1}
% \renewcommand{\ottmv}[1]{#1}
% \renewcommand{\subparagraph}[1]{\vspace{3pt}\noindent{\bf #1}}

% Logical equivalence related macros
\newcommand{\valR}[2]{\mathcal{V}\bra{#1 ; #2}}
\newcommand{\valRR}[1]{\mathcal{V}\bra{#1}}
\newcommand{\eeR}[2]{\mathcal{E}\bra{#1 ; #2}}
\newcommand{\eeRR}[1]{\mathcal{E}\bra{#1}}
\newcommand{\ggR}[2]{\mathcal{G}\bra{#1 ; #2}}

\newcommand{\hll}[2][gray!40]{\colorbox{#1}{#2}}
\newcommand{\hlmath}[2][gray!40]{%
  \colorbox{#1}{$\displaystyle#2$}}


%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

\bibliographystyle{plainurl}% the recommnded bibstyle

\title{The Essence of Nested Composition}

% \titlerunning{Dummy short title}%optional, please use if title is longer than one line

\author{Xuan Bi$^1$}{The University of Hong Kong, Hong Kong, China}{xbi@cs.hku.hk}{}{}%mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.

\author{Bruno C. d. S. Oliveira}{The University of Hong Kong, Hong Kong, China}{bruno@cs.hku.hk}{}{Funded by Hong Kong Research Grant Council projects number 17210617 and 17258816}

\author{Tom Schrijvers}{KU Leuven, Belgium}{tom.schrijvers@cs.kuleuven.be}{}{Funded by The Research Foundation - Flanders}

\authorrunning{X.\,Bi, B.\,C.\,d.\,S.\,Oliveira and T.\,Schrijvers} %mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et. al.'

\Copyright{Xuan Bi, Bruno C. d. S. Oliveira and Tom Schrijvers}%mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/


\subjclass{Software and its engineering $\rightarrow$ Object oriented languages}% mandatory: Please choose ACM 2012 classifications from https://www.acm.org/publications/class-2012 or https://dl.acm.org/ccs/ccs_flat.cfm . E.g., cite as "General and reference $\rightarrow$ General literature" or \ccsdesc[100]{General and reference~General literature}.

\keywords{nested composition, family polymorphism, intersection types, coherence}%mandatory

\category{}%optional, e.g. invited paper

\relatedversion{}%optional, e.g. full version hosted on arXiv, HAL, or other respository/website

\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...

\funding{}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

\acknowledgements{We thank the anonymous reviewers for their helpful comments.}%optional

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{Todd Millstein}
\EventNoEds{1}
\EventLongTitle{32nd European Conference on Object-Oriented Programming (ECOOP 2018)}
\EventShortTitle{ECOOP 2018}
\EventAcronym{ECOOP}
\EventYear{2018}
\EventDate{July 16--21, 2018}
\EventLocation{Amsterdam, Netherlands}
\EventLogo{}
\SeriesVolume{109}
\ArticleNo{22} % “New number” (=<article-no>) goes here!%\nolinenumbers %uncomment to disable line numbering
%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\maketitle

\begin{abstract}
Calculi with \emph{disjoint intersection types} support an introduction form
for intersections called the
\emph{merge operator}, while retaining a \emph{coherent} semantics.
% The interesting feature of such calculi is that they
%retain a coherent semantics, which is known to be hard in the presence
%of the merge operator.
Disjoint intersections types have great potential to
serve as a foundation for powerful, flexible and yet type-safe and
easy to reason OO languages. This paper shows how to significantly
increase the expressive power of disjoint intersection types by
adding support for \emph{nested subtyping and composition}, which
enables simple forms of \emph{family polymorphism} to be expressed in the calculus.
%The motivation for those features is Ernst's \emph{family polymorphism}: the idea
%that inheritance can be extended from a single class, to a whole
%family of classes. Nested subtyping and composition enable simple forms
%of family polymorphism to be expressed, while retaining type-safety
%and coherence.
The extension with nested subtyping and
composition is challenging, for two different reasons. Firstly, the
subtyping relation that supports these features is non-trivial,
especially when it comes to obtaining an algorithmic version. Secondly,
the syntactic method used to prove coherence for previous calculi with disjoint
intersection types is too inflexible, making it hard to
extend those calculi with new features (such as nested subtyping).
We show how to address the first problem by adapting and extending
 the Barendregt, Coppo and Dezani (BCD) subtyping rules for intersections
with records and coercions. A sound and complete algorithmic
system is obtained by using an approach inspired by Pierce's
work. To address the second
problem we replace the syntactic method to prove coherence,
by a semantic proof method based on \emph{logical relations}.
Our work has been fully formalized in Coq, and we have an implementation
of our calculus.
\end{abstract}


% Main meat
\input{sections/introduction.tex}
\input{sections/overview.tex}
\input{sections/typesystem.mng}
\input{sections/coherence.mng}
\input{sections/algorithm.mng}
% \input{sections/discussion.tex}
\input{sections/related.tex}
\input{sections/conclusion.tex}


%%
%% Bibliography
%%

%% Please use bibtex,

\bibliography{paper}


% \ifdefined\submitoption
\newpage
\appendix
\input{appendix.mng}
% \fi

\end{document}
