% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
\usepackage{llncsdoc}
%
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following line
% to display URLs in blue roman font according to Springer's eBook style:
% \renewcommand\UrlFont{\color{blue}\rmfamily}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Load up my personal packages
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage[numbers,sort]{natbib}
\input{packages.tex}
\input{pl-theory.tex}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Hyperlinks
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage[capitalise]{cleveref}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Ott includes
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage{ottalt}
\inputott{ott-rules}
\renewcommand\ottaltinferrule[4]{
  \inferrule*[narrower=0.6,lab=#1,#2]
    {#3}
    {#4}
}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Revision tool
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \newcommand\mynote[3]{\textcolor{#2}{#1: #3}}
% \newcommand\bruno[1]{\mynote{Bruno}{red}{#1}}
% \newcommand\tom[1]{\mynote{Tom}{blue}{#1}}
% \newcommand\ningning[1]{\mynote{Ningning}{orange}{#1}}
% \newcommand\jeremy[1]{\mynote{Jeremy}{purple}{#1}}



\begin{document}
%
\title{Distributive Disjoint Polymorphism for Compositional Programming}
%
\titlerunning{Distributive Disjoint Polymorphism}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Xuan Bi\inst{1} \and
Ningning Xie\inst{1} \and
Bruno C. d. S. Oliveira\inst{1} \and
Tom Schrijvers\inst{2}}
%
\authorrunning{X.\,Bi et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{The University of Hong Kong, Hong Kong, China \\
\email{\{xbi,nnxie,bruno\}@cs.hku.hk} \and
KU Leuven, Leuven, Belgium \\
\email{tom.schrijvers@cs.kuleuven.be}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}

\begin{comment}
Compositionality is a desirable property in programming
designs. Broadly defined, compositionality is the principle that a
system should be built by composing smaller subsystems.
Programming techniques such as \emph{shallow embeddings} of
Domain Specific Languages (DSLs),  \emph{finally tagless} or \emph{object algebras}
are built on the principle of compositionality.
However, programming languages often only support well simple
compositional designs, but language support for more sophisticated
compositional designs is lacking.

In this paper we present a calculus and polymorphic type system with \emph{(disjoint)
intersection types}, called \fnamee,
that supports a broader notion of compositional designs, and enables the
development of highly modular and reusable programs. \fnamee is a
generalization and extension of Alpuim et al. \fname calculus,
which proposed the idea of \emph{disjoint polymorphism}.
The main novelty of \fnamee is a novel subtyping algorithm with
distributivity laws on types. Distributivity plays a fundamental role
in improving compositional designs, by enabling the automatic
composition of multiple operations/interpretations. The main technical
challenge is the proof of coherence for \fnamee as impredicativity makes it
hard to develop a well-founded logical relation for coherence.
However, by restricting the system to predicative instantiations only
we are able to develop a suitable logical relation and show
coherence.
We illustrate the use of the calculus to model highly modular
interpretations of DSLs, that can be smoothly composed with
the merge operator of \fnamee. Furthermore, we provide a detailed
comparison between \emph{distributive disjoint polymorphism} and
\emph{row types}.
\end{comment}

Popular programming techniques such as \emph{shallow embeddings} of Domain
Specific Languages (DSLs), \emph{finally tagless} or \emph{object
  algebras} are built on the principle of \emph{compositionality}.  However,
existing programming languages only support simple compositional designs well,
and have limited support for more sophisticated ones.

This paper presents the \fnamee calculus, which supports
highly modular and compositional designs that
improve on existing techniques. These improvements are due to the
combination of three features: \emph{disjoint intersection
  types} with a \emph{merge operator}; \emph{parametric (disjoint)
  polymorphism}; and \emph{BCD-style distributive subtyping}.  The
main technical challenge is  \fnamee's proof of coherence. A
naive adaptation of ideas used in System F's \emph{parametricity} to
\emph{canonicity} (the logical relation used by \fnamee to prove coherence) results in an ill-founded logical relation. To solve the
problem our canonicity relation employs a different technique based on
immediate substitutions and a restriction to predicative
instantiations. Besides coherence, we show
several other important meta-theoretical results, such as type-safety, 
sound and complete algorithmic subtyping, and
decidability of the type system. Remarkably, unlike 
\fsub's \emph{bounded polymorphism}, disjoint polymorphism
in \fnamee supports decidable type-checking.


\begin{comment}
\fnamee  is a
generalization and extension of Alpuim et al. \fname calculus,
which proposed the idea of \emph{disjoint polymorphism}.
The main novelty of \fnamee is a novel polymorphic subtyping algorithm with
distributivity laws on types. Distributivity plays a fundamental role
in improving compositional designs, by enabling the automatic
composition of multiple operations/interpretations.
The main technical
challenge is the proof of coherence for \fnamee as impredicativity makes it
hard to develop a well-founded logical relation for coherence.
However, by restricting the system to predicative instantiations only
we are able to develop a suitable logical relation and show
coherence.
We illustrate the use of the calculus to model highly modular
interpretations of DSLs, that can be smoothly composed with
the merge operator of \fnamee. Furthermore, we provide a detailed
comparison between \emph{distributive disjoint polymorphism} and \emph{row types}.
\end{comment}

% \keywords{First keyword  \and Second keyword \and Another keyword.}
\end{abstract}
%
%
%
\input{Sources/introduction}
\input{Sources/overview}
\input{Sources/typesystem.mng}
\input{Sources/coherence.mng}
% \input{Sources/row.mng}
\input{Sources/related.mng}
\input{Sources/conclusion}

\bibliographystyle{splncs04}
\bibliography{paper}


\newpage
\appendix
\input{Sources/translation.mng}


\end{document}
